---
title: "Gene set scores computation with NetActivity"
author: 
-   name: Carlos Ruiz Arenas
    affiliation: Centro de Investigación Biomédica en Red de Enfermedades Raras (CIBERER), Barcelona, Spain
    email: carlos.ruiza@upf.edu
date: "`r Sys.Date()`"
package: NetActivity
output: 
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{"Gene set scores computation with NetActivity"}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installation

You can install the development version of NetActivity from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("yocra3/NetActivity")
devtools::install_github("yocra3/NetActivityData")
```

# Introduction

The coordination of multiple genes is required to perform specific biological functions, so grouping gene expression in gene sets provides better biological insights than studying individual genes. We have developed a framework (`NetActivityTrain`) to encode individual gene expression measurements into gene set scores (ref). Our framework is based on sparsely-connected autoencoders. In our framework, each gene set is represented by a neuron in the hidden layer, which is connected to only the gene set genes from the input layer. Nonetheless, to ensure a better representation of the gene sets, all gene sets are connected to all genes in the output layer.

Our framework, `NetActivityTrain`, is implemented in Nextflow, which is useful to train the models but not to apply the models to new data. To overcome this issue, we have implemented the package `r Githubpkg("yocra3/NetActivity")`. `r Githubpkg("yocra3/NetActivity")` enables to easily apply the gene set representations obtained with `NetActivityTrain` to new datasets. We have included two models containing GO Biological processes and KEGG pathways trained with GTEx or TCGA data in `r Githubpkg("yocra3/NetActivityData")`.

# Pre-processing

Computation of gene set scores is performed with `computeGeneSetScores`. The `computeGeneSetScores` functions requires a `r Biocpkg("SummarizedExperiment")` with the gene expression data standardized at gene level. The function `prepareSummarizedExperiment` facilitates the process.

`r Githubpkg("yocra3/NetActivity")` can be applied to RNAseq or microarray data. We will exemplify how to process each dataset independently.

## RNAseq

For this vignette, we will show how to process RNAseq using the `r Biocpkg("airway")` dataset:

```{r, warning = FALSE, message = FALSE}
library(NetActivity)
library(DESeq2)
library(airway)
data(airway)
```

The `airway` dataset contains `r nrow(airway)` genes and `r ncol(airway)` samples. Gene expression data is represented as counts. `r Rpackage("NetActivity")` functions require normalized data to work. We suggest using Variance Stabilizing Transformation from `r Biocpkg("DESeq2")` to normalize gene expression data:

```{r}
ddsSE <- DESeqDataSet(airway, design = ~ cell + dex)
vst <- varianceStabilizingTransformation(ddsSE)
```

Once we have normalized the data, we can proceed with the pre-processing. `prepareSummarizedExperiment` requires two arguments: a `SummarizedExperiment` and a model trained from `NetActivityTrain`. The current version of `r Rpackage("NetActivityData")` includes a model with GO-BP (Biological Processes) terms and KEGG pathways trained with GTEx or with TCGA. In this vignette, we will use the model trained in GTEx:

```{r}
out <- prepareSummarizedExperiment(vst, "gtex_gokegg")
out
```

`prepareSummarizedExperiment` performs two steps. First, it checks whether all genes in the model are present in the input `SummarizedExperiment`. Missing genes are added as a row of 0s. Second, the function standardizes gene expression data by gene. We can compare the genes values before and after `prepareSummarizedExperiment` to see the effect of normalization:

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
```

```{r plot, fig.cap = "Standardization. Effect of standardization on gene expression values. Before represent the gene expression values passed to prepareSummarizedExperiment. After are the values obtained after prepareSummarizedExperiment."}
rbind(assay(vst[1:5, ]) %>%
        data.frame() %>%
        mutate(Gene = rownames(.)) %>%
        gather(Sample, Expression, 1:8) %>%
        mutate(Step = "Before"),
    assay(out[1:5, ]) %>%
        data.frame() %>%
        mutate(Gene = rownames(.)) %>%
        gather(Sample, Expression, 1:8) %>%
        mutate(Step = "After")) %>%
    mutate(Step = factor(Step, levels = c("Before", "After"))) %>%
    ggplot(aes(x = Gene, y = Expression, col = Gene)) +
        geom_boxplot() +
        theme_bw() +
        facet_grid(~ Step, scales = "free") +
        theme(axis.ticks = element_blank(), axis.text.x = element_blank())        
```

In Figure \@ref(fig:plot), we can see that the original gene expression values are normalized values with different medians and dispersions. After standardization, all gene values are centered to a mean of 0 and have more similar distributions.

Next, we can check the values of a gene not present originally in the data:

```{r}
assay(out["ENSG00000278637", ]) 
```

Notice that functions included in `r Rpackage("NetActivity")` require that the `SummarizedExperiment` and the model shared the same annotation. In this example, both objects were annotated using ENSEMBL annotation, so no additional steps were required. In case the genes in the `SummarizedExperiment` use a different annotation, see how to convert them in the next section. Notice that the current models included in `r Rpackage("NetActivityData")` contain the genes using ENSEMBL annotation.

## Microarray data

For this vignette, we will show how to process microarray data using the `r Biocpkg("Fletcher2013a")` dataset:

```{r, message = FALSE, warning = FALSE}
library(limma)
library(Fletcher2013a)
data(Exp1)
Exp1
```

`Exp1` is an `ExpressionSet` containing the information for `r nrow(Exp1)` genes and `r ncol(Exp1)` samples. Genes are named with Affymetrix ids, but the SYMBOL information is available.

The first step is converting the `ExpressionSet` to a `SummarizedExperiment`:

```{r}
SE_fletcher <- SummarizedExperiment(exprs(Exp1), colData = pData(Exp1), rowData = fData(Exp1))
```

Second, we will set the rownames to ENSEMBL. For this, we will first map probe ids to SYMBOL and then to ENSEMBL id. Notice that mapping probes to SYMBOL and then to ENSEMBL might result in probes without annotation or duplicated probes with the same ENSEMBL id. For the sake of simplicity, in this vignette we will remove probes without SYMBOL or with duplicated SYMBOLs based on order. Nonetheless, the users are encouraged to apply other criteria in order to get the final set. 

```{r, warning=FALSE, message = FALSE}
library(AnnotationDbi)
library(org.Hs.eg.db)
```

```{r}
rownames(SE_fletcher) <- rowData(SE_fletcher)$SYMBOL
SE_fletcher <- SE_fletcher[!is.na(rownames(SE_fletcher)), ]
SE_fletcher <- SE_fletcher[!duplicated(rownames(SE_fletcher)), ]

rownames(SE_fletcher) <- mapIds(org.Hs.eg.db,
    keys = rownames(SE_fletcher),
    column = 'ENSEMBL',
    keytype = 'SYMBOL')
SE_fletcher <- SE_fletcher[!is.na(rownames(SE_fletcher)), ]
SE_fletcher <- SE_fletcher[!duplicated(rownames(SE_fletcher)), ]
SE_fletcher
```

Now, we can pass the resulting `SummarizedExperiment` to `prepareSummarizedExperiment`:

```{r}
out_array <- prepareSummarizedExperiment(SE_fletcher, "gtex_gokegg")
out_array
```

# Computing the scores

Once we have pre-processed the data, we are ready to compute the gene set scores. We will continue with the output of the microarray data (`out_array`).

`computeGeneSetScores` requires two arguments: a `SummarizedExperiment` and a model trained from `NetActivityTrain`:

```{r}
scores <- computeGeneSetScores(out_array, "gtex_gokegg")
scores
```

`computeGeneSetScores` returns a `SummarizedExperiment` with the gene set scores for all the gene sets present in the model. The `colData` of the input `SummarizedExperiment` is preserved, so this object can be directly used for downstream analyses. The `rowData` contains annotation of the gene sets, which can be useful for downstream analyses:

```{r}
rowData(scores)
```


# Differential gene set scores analysis

Once the gene set scores are computed, we can test for differential expression using `r Biocpkg("limma")`. We will run a differential analysis due to treatment, adjusted for time points:

```{r}
mod <- model.matrix(~ Treatment + Time, colData(scores))
fit <- lmFit(assay(scores), mod) %>% eBayes()
topTab <- topTable(fit, coef = 2:4, n = Inf)
head(topTab)
```


For easing the interpretation, we can add the gene set full name to the results table:

```{r}
topTab$GeneSetName <- rowData(scores)[rownames(topTab), "Term"]
head(topTab)
```


Many gene sets present high differences due to treatment. We will further explore the top gene set (GO:1990440 or positive regulation of transcription from RNA polymerase II promoter in response to endoplasmic reticulum stress).

Once we have our candidate gene set, we recommend taking a look to the distribution of gene set scores based on our variable of interest:

```{r plotScores, fig.cap = "GO:1990440 activity scores. GO:1990440 presented the most significant difference due to treatment."}
data.frame(Expression = as.vector(assay(scores["GO:1990440", ])),
    Treatment = scores$Treatment) %>%
    ggplot(aes(x = Treatment, y = Expression, col = Treatment)) +
        geom_boxplot() +
        theme_bw() +
    ylab("NetActivity scores")
```


We can clearly see differences in GO:1990440 activation scores between the different treatment groups in Figure \@ref(fig:plotScores). UT samples have the lowest gene set activity scores while E2FGF10 have the highest activation scores. 

Nonetheless, notice that the sign of the gene set score is arbitrary and cannot be directly interpreted. To be able to interpret the scores, we can take a look to the weights used for the computation:

```{r plotWeight, fig.cap = "Weights of GO:1990440 gene set. The figure represents the weights used for computing the GO:1990440 gene set score. Weights are in absolute value to enable an easier comparison of their magnitude. Positive weights are shown in blue and negative in red.", fig.wide = TRUE}
weights <- rowData(scores)["GO:1990440", ]$Weights_SYMBOL[[1]]
data.frame(weight = weights, gene = names(weights)) %>%
    mutate(Direction = ifelse(weight > 0, "Positive", "Negative")) %>%
    ggplot(aes(x = gene, y = abs(weight), fill = Direction)) + 
    geom_bar(stat = "identity") +
    theme_bw() +
    ylab("Weight") +
    xlab("Gene")
```

In Figure \@ref(fig:plotWeight), we can see a comparison between the gene weights used for the gene set score computation.  Genes with larger weights have a higher importance on gene set computation. Thus, ATF6, ATF4, DDIT3, CEBPB and TP53 are the most relevant genes for the gene set computation. As all of them have a positive sign, individuals with higher gene set activity scores for GO:1990440 will have higher expression of these genes. 


```{r}
sessionInfo()
```

